---
title: 'H2O Demo: Human Activity Recognition with Smartphones'
output:
  html_notebook:
    fig_height: 6
    fig_width: 9
    highlight: haddock
    theme: spacelab
  html_document: default
---

## About the dataset

- Recordings of 30 study participants performing activities of daily living
- by UCI Machine Learning
- Reference (Original): https://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones
- Reference (Kaggle): https://www.kaggle.com/uciml/human-activity-recognition-with-smartphones


### Description

The Human Activity Recognition database was built from the recordings of 30 study participants performing activities of daily living (ADL) while carrying a waist-mounted smartphone with embedded inertial sensors. The objective is to classify activities into one of the six activities performed.

### Description of experiment

The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, we captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data.

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain.

### Attribute information

For each record in the dataset the following is provided:

- Triaxial acceleration from the accelerometer (total acceleration) and the estimated body acceleration.
- Triaxial Angular velocity from the gyroscope.
- A 561-feature vector with time and frequency domain variables.
- Its activity label.
- An identifier of the subject who carried out the experiment.

### Relevant papers

- Davide Anguita, Alessandro Ghio, Luca Oneto, Xavier Parra and Jorge L. Reyes-Ortiz. Human Activity Recognition on Smartphones using a Multiclass Hardware-Friendly Support Vector Machine. International Workshop of Ambient Assisted Living (IWAAL 2012). Vitoria-Gasteiz, Spain. Dec 2012

- Davide Anguita, Alessandro Ghio, Luca Oneto, Xavier Parra, Jorge L. Reyes-Ortiz. Energy Efficient Smartphone-Based Activity Recognition using Fixed-Point Arithmetic. Journal of Universal Computer Science. Special Issue in Ambient Assisted Living: Home Care. Volume 19, Issue 9. May 2013

- Davide Anguita, Alessandro Ghio, Luca Oneto, Xavier Parra and Jorge L. Reyes-Ortiz. Human Activity Recognition on Smartphones using a Multiclass Hardware-Friendly Support Vector Machine. 4th International Workshop of Ambient Assited Living, IWAAL 2012, Vitoria-Gasteiz, Spain, December 3-5, 2012. Proceedings. Lecture Notes in Computer Science 2012, pp 216-223.

- Jorge Luis Reyes-Ortiz, Alessandro Ghio, Xavier Parra-Llanas, Davide Anguita, Joan Cabestany, Andreu Català. Human Activity and Motion Disorder Recognition: Towards Smarter Interactive Cognitive Environments. 21st European Symposium on Artificial Neural Networks, Computational Intelligence and Machine Learning, ESANN 2013. Bruges, Belgium 24-26 April 2013.

### Citation

Davide Anguita, Alessandro Ghio, Luca Oneto, Xavier Parra and Jorge L. Reyes-Ortiz. A Public Domain Dataset for Human Activity Recognition Using Smartphones. 21st European Symposium on Artificial Neural Networks, Computational Intelligence and Machine Learning, ESANN 2013. Bruges, Belgium 24-26 April 2013.

<hr>

<br>

## Step 1 - Start and connect to a H2O cluster (JVM)

```{r}
# Pre-load all R packages
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(h2o))
suppressPackageStartupMessages(library(plotly))
```

```{r}
# Start and connect to a H2O cluster (JVM)
h2o.init(nthreads = -1)
h2o.no_progress() # disable progress bar in notebbok
```

<br>

## Step 2 - Importing datasets into H2O

```{r}
# Import pre-processed datasets

# locally (if you have the datasets in the 'data' sub-folder)
hex_train <- h2o.importFile("./data/train.csv.gz")
hex_test <- h2o.importFile("./data/test.csv.gz")

# or directly from the web (github)
# hex_train <- h2o.importFile("https://github.com/woobe/h2o_demos/blob/master/human_activitiy_recognition_with_smartphones/data/train.csv.gz?raw=true")
# hex_test <- h2o.importFile("https://github.com/woobe/h2o_demos/blob/master/human_activitiy_recognition_with_smartphones/data/test.csv.gz?raw=true")
```

<br>

## Step 3 - Exploratory Analysis

```{r}
# Dimensions
# 'Train' dataset has 7352 rows and 562 columns
# 'Test' dataset has 2947 rows and 562 columns
dim(hex_train)
dim(hex_test)
```

```{r}
# First few records
# First column is the label 'activity'
# Rest of the columns (V1 to V561) are sensors data
head(hex_train)
head(hex_test)
```


```{r}
# Look at 'activity' column
# Six classes (Carinality = 6)
# No missing value
h2o.describe(hex_train$activity)
h2o.describe(hex_test$activity)
```


```{r}
# Extract 'activity' columns for other graphics packages in R
d_activity_train <- as.data.frame(hex_train$activity)
d_activity_test <- as.data.frame(hex_test$activity)

# Count acitivity 
d_freq_train <- as.data.frame(table(d_activity_train))
d_freq_test <- as.data.frame(table(d_activity_test))
d_freq <- merge(d_freq_train, d_freq_test, by.x = "d_activity_train", by.y = "d_activity_test", sort = FALSE)
colnames(d_freq) <- c("activity", "freq_train", "freq_test")
d_freq
```

```{r, fig.width = 9, fig.height = 6}
# Visualize 'activity' in both 'train' and 'test'
p <- plot_ly(d_freq, x = ~activity, y = ~freq_train, type = 'bar', name = 'Frequency (Train)') %>%
  add_trace(y = ~freq_test, name = 'Frequency (Test)') %>%
  layout(title = "Activities in 'Train' and 'Test' Dataset") %>%
  layout(yaxis = list(title = 'Count'), xaxis = list(title = "")) %>%
  layout(margin = list(b = 90)) %>%
  layout(barmode = "group")
p
```

```{r}
# Look at relationship between sensor data V1 and activity
d_v1 <- data.frame(V1_train = as.data.frame(hex_train$V1), activity = as.data.frame(hex_train$activity))
head(d_v1)
```

```{r, fig.width = 9, fig.height = 6}
p <- plot_ly(d_v1, y = ~V1, color = ~activity, type = "box") %>%
     layout(title = "Relationship between Sensor Data V1 and Activities") %>%
     layout(yaxis = list(title = 'Sensor Data V1'), xaxis = list(title = "")) %>%
     layout(margin = list(b = 90))
p
```

```{r}
# Principal Component Analysis
# 95% of variance in original data captured by first five principal components
suppressWarnings(
    model_pca <- h2o.prcomp(training_frame = hex_train, 
                        x = 2:562, 
                        model_id = "h2o_pca",
                        k = 5)    
)
model_pca    
```

```{r}
# Visualize principle components with activity labels
d_pca <- as.data.frame(h2o.predict(model_pca, hex_train))
d_pca <- data.frame(d_pca, as.data.frame(hex_train$activity))
head(d_pca)
```

```{r, fig.width = 9, fig.height = 6}
p <- plot_ly(data = d_pca, x = ~PC2, y = ~PC3, color = ~activity, 
             type = "scatter", mode = "markers", marker = list(size = 3)) %>%
     layout(title = "Visualizing Principle Components")
p
```

From the graph above, we can see that:

- it could be difficult to distinguish between **Standing** and **Sitting** as there are large overlaps in their sensor data.
- **Laying** has its own cluster so it should be easy to classify.
- **Walking**, **Walking Upstairs** and **Walking Downstairs** are understandably closer to each other yet they are quite different to **Sitting**, **Standing** and **Laying**.


<br>

## Step 4 - Build and evalutate a predictive model using H2O's Gradient Boosting Machine (GBM) algorithm

```{r}
# Define target and features for model training
target <- "activity"
features <- setdiff(colnames(hex_train), target) # i.e. using the records of all 561 sensors
```

```{r}
# Build a GBM model
model <- h2o.gbm(x = features,
                 y = target,
                 training_frame = hex_train,                 
                 model_id = "h2o_gbm",
                 ntrees = 500,
                 learn_rate = 0.05,
                 learn_rate_annealing = 0.999,
                 max_depth = 7,
                 sample_rate = 0.9,
                 col_sample_rate = 0.9,
                 nfolds = 3,
                 fold_assignment = "Stratified",
                 stopping_metric = "logloss",
                 stopping_rounds = 5,
                 score_tree_interval = 10,
                 seed = 1234)
```

```{r}
# Print out model summary
model
```


<br>

## Step 5 - Make and evalutate predictions

```{r}
# Make predictions
yhat_test <- h2o.predict(model, hex_test)
head(yhat_test)
```

```{r}
# Evaluate predictions
h2o.performance(model, newdata = hex_test)
```

As expected:
- It is easy to classify **Laying**
- It is difficult to distinguish between **Sitting** and **Standing**



<br>

## Step 6 - Export the GBM model as MOJO for Shiny applications

```{r}
# Not Run
# Showing the syntax for now
# h2o.saveMojo(model, path = "")
```

<br>

TO DO:

- Variable importance graph/table
- Scoring history
- Compressed data size in JVM


